---
title: "Max is not Normal"
author: "Peter Rabinovitch"
date: "`r Sys.time()`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<details><summary>Code</summary>

```{r, warning=FALSE,error=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
```

</details>

# Introduction

Recently I have seen a number of systems that attempt to detect anomalies, whether it be in the number of transactions, the time to download a web page, or the sales of a product.

Of course this is a good thing to do - to make sure that if anything is out of whack that you are alerted and can investigate and take appropriate action.

A common theme among these systems if the trade-off between false alarms and missing true anomalies. One approach that seems ubiquitous is that of comparing the current value to historical values (perhaps using covariates) and alerting of the current value falls outside some interval. Usually this is based on (some variant of) a t-test, and looks at the mean and (typically) a 95% confidence interval. All good.

What isn't so good is applying the same statistical approach to the maximum (or minimum). Here we expand on the title of this post: the maximum is not normally (Gaussian) distributed, and so systems that look at the average maximum over history and compare it to the current maximum and alert if the current values falls outside a 95% confidence interval, determined by the mean of the max values +/+ 1.96 * the standard deviation of the historical maxima, is not correct.

# Illustration

Here we show what goes wrong.

<details><summary>Code</summary>

```{r}
set.seed(2024)
d <- 365 # days
hpd <- 24 # hours per day
nph <- 60 # number of data points per hour
mu <- 100 # mean of the data points
sigma <- 20 # sd of the data points

df <- tibble(x = rnorm(d * hpd * nph, mu, sigma)) %>%
  rowid_to_column() %>%
  mutate(h = ceiling(rowid / nph)) # sample of the data

dg <- df %>% # look at it on an hourly basis
  group_by(h) %>%
  summarize(M = max(x), s = sum(x), m = mean(x), v = var(x), .groups = "drop")

# plot the data & fom's
al <- 0.1

p1 <- dg %>% ggplot(aes(x = h, y = m)) +
  geom_point(alpha = al) +
  theme_minimal() +
  labs(x = "", y = "Hourly Average") +
  scale_x_continuous(label = scales::comma) +
  scale_y_continuous(label = scales::comma)

p2 <- dg %>% ggplot(aes(x = h, y = M)) +
  geom_point(alpha = al) +
  theme_minimal() +
  labs(x = "Hour", y = "Hourly Max") +
  scale_x_continuous(label = scales::comma) +
  scale_y_continuous(label = scales::comma)
```

</details>

```{r,echo=FALSE}
p1 / p2
```

We plot the hourly average of some figure of merit (fom), as well as the maximum of the fom, by hour. 

Next we show the distributions of these values, along with their best fitting Gaussian distribution in red.

<details><summary>Code</summary>

```{r}
# plot the histograms & best Gaussian to the data
nb <- 200

p1 <- ggplot() +
  geom_histogram(data = dg, aes(x = m, y = after_stat(density)), bins = nb) +
  geom_function(fun = dnorm, args = list(mean = mean(dg$m), sd = sd(dg$m)), colour = "red") +
  theme_minimal() +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Hourly Avarage", y = "")

p2 <- ggplot() +
  geom_histogram(data = dg, aes(x = M, y = after_stat(density)), bins = nb) +
  geom_function(fun = dnorm, args = list(mean = mean(dg$M), sd = sd(dg$M)), colour = "red") +
  theme_minimal() +
  labs(x = "Hourly Max", y = "")
```

</details>

```{r, echo=FALSE}
p1 / p2
```

The hourly average is very well approximated by a Gaussian, but the hourly maximum is not as well fitted. So, of course, any alarm based on a rule designed for a Gaussian distribution would not be correct for the maximum, because it is not Gaussian distributed.

# Conclusion

Don't blindly apply formulae that you don't really understand. Get out your stats text (or Wikipedia) or talk to someone who knows more.

# Appendices

<details><summary>References</summary>

[Wikipedia post on Extreme Value Theory](https://en.wikipedia.org/wiki/Extreme_value_theory)

</details>

<details><summary>SessionInfo</summary>

```{r}
sessionInfo()
```

</details>






